---
title: '2009-2013 LA Traffic-related Fatalities & Attributes: A Surface Level Examination'
author: Sally Rong, Elena Castellanos
date: 08/06/2020
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---

[Presentation Slides](https://docs.google.com/presentation/d/1yHIeG8eF5opxm5KL7_Ot6Ze8ppijTDbmRJoZ1XdJO2A/edit?usp=sharing)

[YouTube Presentation](https://youtu.be/baXxOzjo_-I)

[SWITRS 2009-2013 Collisions Dataset](https://geohub.lacity.org/datasets/ladot::collisions-2009-2013-switrs?geometry=-149.246%2C-12.366%2C30.578%2C43.974)

[Victims Table](https://geohub.lacity.org/datasets/ladot::victim-tables-collisions-2009-2013-switrs)

[Party Table](https://geohub.lacity.org/datasets/ladot::party-tables-collisions-2009-2013-switrs)

[Codebook](https://tims.berkeley.edu/help/SWITRS.php#Collision_Level)

The codebook provides analysis on 3 levels: the party level, victim level, and collision level.

#rm(list=ls()) 
#setwd("~/Desktop/Data Science") 
#install.packages('tidyverse')

### Load the data
```{r}
library(tidyverse)
collisions2 <- read_csv('Collisions_2009-2013_(SWITRS).csv')
party_table <- read_csv('Party_Tables_-_Collisions_2009-2013_(SWITRS).csv')
victim_table <- read_csv('Victim_Tables_-_Collisions_2009-2013_(SWITRS).csv')
```

### PARTY-VICTIM relationship

#### Merge party_table & victim_table, by CASE ID and any number of party records that can be associated w/ a collision
```{r}
party_victim <- left_join(x=party_table, y=victim_table, by= c("CASE_ID", "PARTY_NUMBER"))
dim(party_victim)
```

### COLLISION-PARTY relationship

#### For this join, we faced difficulting tying the two datasets by CASE_ID. We have ran this problem through with Professor Raja too.
```{r}
#collision_party <- left_join(x=collisions2, y=party_table, by= c("CASE_ID"))
```

```{r}
df <- party_victim %>%
  select(PARTY_AGE, PARTY_SEX, VICTIM_DEGREE_OF_INJURY, VICTIM_SAFETY_EQUIP_1, VICTIM_ROLE, VICTIM_AGE, VEHICLE_YEAR, VICTIM_SEX, STWD_VEHICLE_TYPE, VEHICLE_MAKE)
dim(df)
```

#### We first looked only at fatal accidents.
```{r}
fatal <- filter(df, VICTIM_DEGREE_OF_INJURY == 4)
dim(fatal)
```


```{r}
fatal %>%
  group_by(VICTIM_SAFETY_EQUIP_1) %>%
  summarize(num = n()) %>%
  arrange(desc(num)) %>%
  mutate(percentage = num/277934)
```
#### It appears that in accidents that resulted in the victim killed,  about 46.3% of those accidents involved the air bag not deployed ('M' above). 


##### Here's a graph showing the Victim Safety Equipment used in fatal traffic accidents. Most frequently occuring is the air bag not deployed (M).
```{r}
fatal %>%
  ggplot(mapping = aes(x=VICTIM_SAFETY_EQUIP_1)) +
  geom_histogram(
    stat = 'count') +
    labs(y = 'Frequency Count',
          x= 'Victim Safety Equipment',
        title = 'Traffic Fatalies and Saftey Equipment Involved')
```

```{r}
no_air_bag_fatal <- filter(df, VICTIM_SAFETY_EQUIP_1 == 'M', VICTIM_DEGREE_OF_INJURY == 4)
dim(no_air_bag_fatal)
```
#### 128730/277934 of traffic accidents in which victim was killed involved air bag not deployed


#### We then looked at the most frequent ages of parties involved in fatal accidents in which the air bag was not deployed.
```{r}
no_air_bag_fatal %>%
  filter(PARTY_AGE != 998) %>%
  group_by(PARTY_AGE) %>%
  summarize(number_parties_in_fatal_accidents = n()) %>%
  arrange(desc(number_parties_in_fatal_accidents))
```
#### Younger parties in their twenties are more frequent in fatal traffic accidents of no air bag deployment. 

```{r}
no_air_bag_fatal %>%
  filter(PARTY_AGE != 998 ) %>%
  ggplot(mapping = aes(x=PARTY_AGE, fill = PARTY_SEX, color= PARTY_SEX)) +
  geom_histogram(
    stat = "count",
    bin = 0.5) +
  labs(y = 'Frequency Count',
       x= 'Party Age',
        title = 'Frequency of Parties Involved in Fatal Traffic Accidents of No Air Bag Deployment by Age')
```

```{r}
filter (no_air_bag_fatal) %>%
  group_by(VEHICLE_MAKE) %>%
  summarize(num = n()) %>%
  arrange(desc(num))
```
#### Of the fatal accidents without the airbag deployed, these are the top car makes.

```{r}
no_air_bag_fatal %>%
  filter( VEHICLE_MAKE %in% c('TOYOTA', 'HONDA', 'FORD',  'NISSAN', 'CHEVROLET')) %>%
  ggplot(mapping = aes(x=VEHICLE_MAKE, fill = VEHICLE_YEAR, color= VEHICLE_YEAR)) +
  geom_histogram(
    stat = 'count') +
    theme_light() +
    labs(y = 'Frequency Count',
          x= 'Car Makes',
        title = 'Top 5 Car Makes in Fatal No-Airbag-Deployed Accidents')
```
#### Interestingly, some car companies had to recall their cars for defective airbags (more info in the links below). The companies with most cars with the defective Takata airbags were also Honda, Ford, and Toyota. 
https://thomasjhenrylaw.com/blog/product-liability/takata-airbag-recall-a-complete-list-of-affected-vehicles/
https://www.consumerreports.org/car-recalls-defects/takata-airbag-recall-everything-you-need-to-know/]
 https://www.newsweek.com/airbag-recall-list-car-models-toyota-ford-honda-bmw-nissan-takata-1483468

```{r}
no_air_bag_fatal %>%
  group_by(VEHICLE_YEAR) %>%
  summarize(num = n()) %>%
  arrange(desc(num))
```
#### Of Vehicles involved in fatal accidents of no air bag deployment, these are the top car years of vehicles involved. This dataset doesn't include information on specific car models, though the articles do point out the top car make and models involved in the Takata air bag recall. 

-------------------------

#### Now we will examine fatal accidents including bicyclists only. 
```{r}
bicyclist_fatal <- filter(df, VICTIM_ROLE == 4 & VICTIM_DEGREE_OF_INJURY == 4)
dim(bicyclist_fatal)
```

```{r}
bicyclist_fatal %>%
  filter(VICTIM_SAFETY_EQUIP_1 != '-' ) %>%
  group_by(VICTIM_SAFETY_EQUIP_1) %>%
  summarize(num = n()) %>%
  arrange(desc(num))
```
#### Most frequent was P - not required. We didn't evaluate that, since the codebook wasn't clear on what that means. We looked at V instead, the 2nd most frequent - 'Driver, Motorcycle Helmet not Used.' We also included X - 'Passenger, motorcycle, Helmet not used' for a more robust look at cases that the helmet was not used. 

```{r}
bicyclist_fatal %>%
  filter(VICTIM_SAFETY_EQUIP_1 %in% c('V', 'X')) %>%
  group_by(VICTIM_AGE) %>%
  summarize(num = n()) %>%
  arrange(desc(num))
```
### Again, it seems bicyclists in at very young ages, 18-20 years, appear the most frequent in these fatal accidents in which helmets aren't used. Bicyclists over the age of 18 are not required to wear helmets. Perhaps deeper analysis may provide information whether there is correlation between these two factors.
https://www.bicyclelaw.com/bicycle-laws/california-bicycle-laws/california-bicycle-helmet-law/ 

```{r}
bicyclist_fatal %>%
  filter(VICTIM_SAFETY_EQUIP_1 %in% c('V', 'X') & VICTIM_AGE != 998) %>%
  ggplot(mapping = aes(x=VICTIM_AGE, fill = VICTIM_SEX, color= VICTIM_SEX)) +
  geom_histogram(
    stat = "count",
    bin = 0.5) +
  labs(y = 'Frequency Count',
       x= 'Bicyclist Age',
        title = 'Frequency of Bicyclist Victims Involved in Fatal Traffic Accidents without Helmet')
```



